---
title: "Homework 4"
author: "Kyle Frankel"
date: 'Assigned: Feb 10, 2021'
output: html_document
---

##### This homework is due by **6:00 PM on Wednesday, Feb 24**.  

##### To complete this assignment, follow these steps:

1. Download the `homework4.Rmd` file from Moodle

2. Open `homework4.Rmd` in RStudio.

3. Replace the "Your Name Here" text in the `author:` field with your own name.

4. Supply your solutions to the homework by editing `homework4.Rmd`.

5. When you have completed the homework and have **checked** that your code both runs in the Console and knits correctly when you click `Knit HTML`, rename the R Markdown file to `homework4_YourNameHere.Rmd`, and submit on Moodle  (YourNameHere should be changed to your own name.)

##### Homework tips:

1. Recall the following useful RStudio hotkeys.

Keystroke | Description
----------|-------------
`<tab>` | Autocompletes commands and filenames, and lists arguments for functions.
`<up>` | Cycles through previous commands in the console prompt
`<ctrl-up>` | Lists history of previous commands matching an unfinished one
`<ctrl-enter>` | Runs current line from source window to Console. Good for trying things out ideas from a source file.
`<ESC>` | Aborts an unfinished command and get out of the + prompt

**Note**: Shown above are the Windows/Linux keys.  For Mac OS X, the `<ctrl>` key should be substituted with the `<command>` (&#8984;) key.

2. Instead of sending code line-by-line with `<ctrl-enter>`, you can send entire code chunks, and even run all of the code chunks in your .Rmd file. Look under the <Chunks> menu of the Source panel.

3. Run your code in the Console and Knit HTML frequently to check for errors.

4. You may find it easier to solve a problem by interacting only with the Console at first, or by creating a separate `.R` source file that contains only R code and no Markdown.

### Problem 1: Linear regression with bikeshare data

```{r}
library(ggplot2)
library(plyr)
```

For this problem we'll be working with two years of bikeshare data from the Capital Bikeshare system in Washington DC.  The dataset contains daily bikeshare counts, along with daily measurements on environmental and seasonal information that may affect the bikesharing.  

Let's start by loading the data.

```{r}
bikes <- read.csv("bikes.csv", header = TRUE)

# Transform temp and atemp to degrees C instead of [0,1] scale
# Transform humidity to %
# Transform wind speed (multiply by 67, the normalizing value)

bikes <- transform(bikes,
                   temp = 47 * temp - 8,
                   atemp = 66 * atemp - 16,
                   hum = 100 * hum,
                   windspeed = 67 * windspeed)
```

Here's information on what the variables mean.

  - instant: record index
	- dteday : date
	- season : season (1:Spring, 2:Summer, 3:Fall, 4:Winter)
	- yr : year (0: 2011, 1:2012)
	- mnth : month ( 1 to 12)
	- hr : hour (0 to 23)
	- holiday : weather day is holiday or not (extracted from http://dchr.dc.gov/page/holiday-schedule)
	- weekday : day of the week
	- workingday : if day is neither weekend nor holiday is 1, otherwise is 0.
	+ weathersit : 
		- 1: Clear, Few clouds, Partly cloudy, Partly cloudy
		- 2: Mist + Cloudy, Mist + Broken clouds, Mist + Few clouds, Mist
		- 3: Light Snow, Light Rain + Thunderstorm + Scattered clouds, Light Rain + Scattered clouds
		- 4: Heavy Rain + Ice Pallets + Thunderstorm + Mist, Snow + Fog
	- temp : Temperature in Celsius. 
	- atemp: Feeling temperature in Celsius. 
	- hum: Normalized humidity. The values are divided to 100 (max)
	- windspeed: Normalized wind speed. The values are divided to 67 (max)
	- casual: count of casual users
	- registered: count of registered users
	- cnt: count of total rental bikes including both casual and registered

**(a)** Season: Factor or numeric?

Consider the following two regression models.

```{r}
bikes.lm1 <- lm(cnt ~ yr + temp + windspeed + season, data = bikes)

bikes.lm2 <- lm(cnt ~ yr + temp + windspeed + as.factor(season), data = bikes)
```

**What is the difference between these two models?**
```{r}
bikes.lm1
```


```{r}
bikes.lm2
```

*Answer here*
The difference between these two models is that the second model separates seasons 2-4 to be their own variable. In the first model, only season 1 (Spring) is inclduded. However with the as.factor code, the second model separates the seasons. In the second model, it includes season 2, 3, and 4, and compares it to the baseline from model 1 which only includes season 1. 

**(b)** What is the interpretation the coefficient(s) of `season` in each of the models in part (a)?

```{r}
summary(bikes.lm1)
```

```{r}
summary(bikes.lm2)
```

*Answer here*
The coefficient in Linear Model for season (Spring) 1 is 370.38, which means that bike sharing increases by 370.38 when the season is Spring.The coefficients for the other seasons are: Season 2 (Summer) 1082.04, Season 3 (Fall) 807.75, Season 3 (Winter) 1364.63. This means that for each respective season, the amount of bike shares will increase by their respective amounts

**(c)**  Which model makes more sense?
*Answer here*
The second model seems to be the better model, since we can separate the seasons from each other. I cannot think of a time when we would only be interested in looking at the Spring semester. 

**(d)**  Using `ggplot2` graphics, construct a scatterplot of `cnt` (bikeshare count) across `mnth` (month of the year).  Describe what you see.  Would a linear model be a good way of modeling how bikeshare count varies with month?  

```{r}
ggplot(data = bikes, aes(x = mnth, y = cnt)) + geom_point(aes(color = factor(season))) + geom_smooth (method = "lm")
# It seems like the colder months have less bikeshare counts, which makes sense. A linear model probably would not work since the numbers go up and down depending on the season, making it a non-linear relationship.
```

**(e)**  Consider the following three models.  Figures of the model fits are shown to help you interpret the models.

```{r}
# Fit and plot first model
bikes.lm3 <- lm(cnt ~ mnth, data = bikes)

qplot(data = bikes, x = mnth, y = cnt) + stat_smooth(method = "lm")
```

```{r}
# Fit and plot second model
bikes.lm4 <- lm(cnt ~ as.factor(mnth), data = bikes)

# Construct data frame that has fitted values for each month
mnth.df <- data.frame(mnth = unique(bikes$mnth))
df.lm4 <- data.frame(mnth = unique(bikes$mnth),
                     fitted = predict(bikes.lm4, newdata = mnth.df))

# Red points are the fitted values from the model
qplot(data = bikes, x = mnth, y = cnt, alpha = 0.5) + 
  guides(alpha = FALSE) + 
  geom_point(data = df.lm4, aes(x = mnth, y = fitted), colour = I("red"),
             alpha = 1, size = I(5))
```


```{r}
# Fit and plot third model
bikes.lm5 <- lm(cnt ~ mnth + I(mnth^2), data = bikes)

qplot(data = bikes, x = mnth, y = cnt) + 
  stat_smooth(method = "lm", formula = y ~ x + I(x^2))
```

**What are the differences between the models?  How many coefficients are used to model the relationship between bikeshare count and month in each of the models?**
*Answer here*
Each model is different. The first model is linear, trying to explain a relationship between month and bike share counts, similar to what we saw earlier, this model only one coefficient: month. The second model however, uses 12 different coefficients for each of the different months and predicts the amount of bike shares for each month. It shows the prediction with the red dot. The last model goes back to one coefficients: month, but squares it, leading to an exponential model. 

**(f)** Use the `plot()` function to produce diagnostic plots for `bikes.lm3`.  Do you observe any problems?  

```{r}
# Edit me
Diag.bikes.lm3 <- plot(bikes.lm3) 
# The main issue I see is that on the Normal Q-Q plot, initially, the data fits well, but towards the end, it is no longer fitting the line, showing that the data is not normally distributed.
```

**(g)** Produce diagnostic plots for `bikes.lm4` and `bikes.lm5`.  Are any of the problems you identified in part (f) resolved?  Do any remain?  Explain.

```{r}
# As can be seen below, the Normal Q-Q plot still shows the same issues as it did earlier. However when looking back at lm3's residuals vs fitted, we can see a curve, compared to lm4 & lm5's straighter lines. This is good, because it shows how the model fits better.
Diag.bikes.lm4 <- plot(bikes.lm4) 
```

```{r}
Diag.bikes.lm5 <- plot(bikes.lm5) 
```



### Problem 2: Interpreting and testing linear models

This problem once again uses the `bikes` data.

**(a)** Use the `transform` and `mapvalues` functions to map the values of the `season` and `weathersit` variables to something more interpretable.  Your new `weathersit` variable should have levels: Clear, Cloud, Light.precip, Heavy.precip

```{r}
# Edit me
bikes <- transform(bikes,season = mapvalues(season, from = c("1", "2", "3", "4"), to = c("Spring","Summer", "Fall","Winter" )))
bikes <- transform(bikes, weathersit = mapvalues(weathersit, from = c("1", "2", "3", "4"), to = c("Clear","Cloud", "Light.precip","Heavy.precip" )))
head(bikes,3)
```

**(b)** Fit a linear regression model with `cnt` as the outcome, and season, workingday, weathersit, temp, atemp and windspeed as covariates.  Use the `summary` function to print a summary of the model showing the model coefficients.

Note that you may wish to code the workingday variable as a factor, though this will not change the estimated coefficient or its interpretation (workingday is already a 0-1 indicator).


```{r}
# Edit me
bikes_lm <- lm(cnt ~ season + as.factor(workingday) + weathersit + temp + atemp + windspeed, data = bikes)

summary(bikes_lm)
```

**(c)** How do you interpret the coefficient of `workingday` in the model?  What is its p-value?  Is it statistically significant?

```{r}
# Edit me
# When working day is 1, bike sharing counts will increase by 177.861 bikesharing counts. P-value is 0.093, it is only significant at the .1 level, but we usually use .05, so it is not statistically significant.
```

**(d)** How do you interpret the coefficient corresponding to Light.precip weather situation?  What is its p-value?  Is it statistically significant?

```{r}
# Edit me
# When Light.precip is 1, bike sharing counts will decrease by 2738.769. P-value is 2e-16, meaning it is highly statiscally significant. 
```

**(e)** Use the `pairs` function to construct a pairs plot of temp, atemp, hum and windspeed.  The bottom panel of your plot should show correlations (see example in Lecture 5)


```{r}
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
  usr <- par("usr"); on.exit(par(usr))
  par(usr = c(0, 1, 0, 1))
  r <- abs(cor(x, y))
  txt <- format(c(r, 0.123456789), digits = digits) [1]
  txt <- paste0(prefix, txt)
  if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
  text(0.5, 0.5, txt, cex = pmax(1, cex.cor *r))
}
panel.cor
pairs(~ temp + atemp + hum + windspeed, lower.panel = panel.cor, data = bikes )
```

**Do you observe any strong colinearities between the variables?**

```{r}
# There is a .99 correlation between temp and atemp, showing they are strong colinearities.
```

**(f)** Use the `update` function to update your linear model by removing the `temp` variable.  Has the coefficient of `atemp` changed?  Has it changed a lot?  Explain.

```{r}
summary(bikes_lm)
# Yes, it has changed a lot. It increased from -4.559 to 90.843. It also went from not being statistically signifacnt to becoming significant at the .001 level. This came from eliminating the temp variable, thus putting all the emphasis on atemp. 
```

```{r}
bikes_lm_notemp <- update(bikes_lm, .~. -temp)
summary(bikes_lm_notemp)
```


**(g)** How do you interpret the coefficient of `atemp` in the model from part (f)?

A one degree increase in atemp leads to a 90.843 increase in bike sharing counts. 


**(h)** Use the `anova()` function on your model from part (f) to assess whether `weathersit` is a statistically significant factor in the model.  Interpret your findings.  

```{r}
# Edit me
anova(update(bikes_lm_notemp,.~.-weathersit), bikes_lm_notemp, test = "Chisq")
# Since the p-value is 2.2e-16, weathersit is a highly statistically significant variable.
```


